---
title: "Analysis script: Movement patterns of equines into the African horse sickness controlled area of South Africa"
author: "John Duncan Grewar, Michael John Koning, Debra Carter, Lesley S. van Helden, Kerry Loxley, Camilla T. Weyer"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Section 1 - Creating the routes and database

The following libraries are required for this section

```{r section1Required, eval=FALSE}
#Language: R
library(osrm)
library(sf)
library(RPostgreSQL)
```

[Open Street Map routes](https://project-osrm.org/) are created using the prepared data set.

```{r, eval=FALSE}
#Language: R
routing <- read.csv("data/2022Horses.csv")
#An empty routes data frame to absorb the returning routes from OSM 
routes = data.frame()

for (i in 1:nrow(routing)){
  temproute = merge(osrmRoute(c(routing[i,]$src_longitude, routing[i,]$src_latitude), 
                              c(routing[i,]$dst_longitude, routing[i,]$dst_latitude), 
                              overview = 'full'),
                    routing[i,])
  routes = rbind(routes, temproute)
}

routes$movementdate = as.Date(routes$movementdate)
```

For management of the data use a [PostgreSQL](https://www.postgresql.org/download/) database. We use [pgAdmin](https://www.pgadmin.org/) to interact with our PostgreSQL databases and this is likely to have been installed, particularly if you use [EDB](https://www.enterprisedb.com/downloads/postgres-postgresql-downloads) installers. The following `R` script will create a database `horsemovementszaf` in the local environment.

```{r, eval=FALSE}
#Language: R
con <- dbConnect(PostgreSQL(), 
                 dbname = "postgres", 
                 user = "yourusernamehere", 
                 password = "yourpasswordhere", 
                 host = "localhost", 
                 port = 5432)

dbSendQuery(con, "CREATE DATABASE horsemovementszaf")
dbDisconnect(con)
```

In your local database created above please make sure you have the the `postgis` and `postgis_raster` extensions activated for working with spatial objects and raster files. These extensions can be installed after opening a query in your `horsemovementszaf` database.

```{sql, eval=FALSE}
--Language: SQL
CREATE EXTENSION postgis;
CREATE EXTENSION postgis_raster;
```

Write the newly created spatial routing file to the database

```{r, eval=FALSE}
#Language: R
con <- dbConnect(
  PostgreSQL(),
  dbname = 'horsemovementszaf',
  host = 'localhost',
  port = 5432,
  user = 'yourusernamehere',
  password = 'yourpasswordhere'
)

#Write the sf object into the database
table_name <- "routing2022"

st_write(
  routes, 
  con, 
  layer = table_name, 
  driver = "PostgreSQL",
  append = FALSE # this overwrites the existing table if it is there
)

dbDisconnect(con)
```

## Section 2 - Rasterise movement routes for aggregation at annual and monthly intervals

### Import routes data with added month variable from database

```{r, eval=FALSE}
#Language: R
library(terra)
library(rpostgis)
library(raster)
library(RPostgreSQL)

con <- dbConnect(
  PostgreSQL(),
  dbname = 'horsemovementszaf',
  host = 'localhost',
  port = 5432,
  user = 'yourusernamehere',
  password = 'yourpasswordhere'
)

#Query the routes but include a month number to evaluate at monthly intervals
routes = pgGetGeom(con, 
query = "
SELECT id, 
EXTRACT(MONTH FROM movementdate) AS month_number, 
totalequids,
geometry AS geom
FROM routing2022;"
)

dbDisconnect(con)
```

### Rasterise and write rasters to database

```{r, eval=FALSE}
#Language: R

#Create an empty raster bound by South African borders
SA_boundary <- terra::vect("./data/rsa.shp")
r <- terra::rast(SA_boundary, res=0.0866)

#Open database connection
con <- dbConnect(
  PostgreSQL(),
  dbname = 'horsemovementszaf',
  host = 'localhost',
  port = 5432,
  user = 'yourusernamehere',
  password = 'yourpasswordhere'
)

# Loop over each month_number
# Establish raster on the totalequids field for each month
# Write each monthly raster to PostgreSQL
for  (i in 1:12){
  pgWriteRast(con, paste0("raster", i), raster::raster((
    terra::rasterize(terra::vect(routes[which(routes$month_number == i), ]),
                     r,
                     field = "totalequids",  
                     sum = TRUE)
  )), append = FALSE)
}

# Write the full year raster as well
pgWriteRast(con, "raster2022", raster::raster(
  (terra::rasterize(
    terra::vect(routes),
    r, 
    field = "totalequids",  
    sum = TRUE
    )
   )
  ), append = FALSE)

#Close database connection
dbDisconnect(con)
```

## Section 3 - Establish risk proportions for routes

The proportion of each route that is made up of different AHS risk classes is defined in short through the following process:

-   Establish raster of the AHS risk across the country on a daily basis

-   Extract the raster values for each pixel along each route on the applicable day

-   Establish the proportion each route has for each of the risk classes, allocating zero to risk classes that did not intersect each route

-   This process is repeated for aggregation at a monthly basis

### Risk proportions for the entire year

```{r, eval=FALSE}
#Language: R

library(sf)
library(terra)
library(data.table)
library(ggplot2)
library(gridExtra)

#source data
zaf <- terra::vect("./data/rsa.shp") #South African boundary
r <- terra::rast(zaf, res=0.0866)
svareas = st_read("./data/svareas.shp") #State vet areas of South Africa
asd <- read.csv("data/2022asd.csv") # Area status declaration per State Vet area per day of the year

# use routes imported from database in previous section
# convert to a sf object
routes = sf::st_as_sf(routes)

#create empty vectors to put proportion values into
low_propVect <- c()
high_propVect <- c()
partial_propVect <- c()
unknown_propVect <- c()

#loop over unique movement dates
for(i in as.character(unique(routes$movementdate))){ # 1. for every movement date 
  #Create temporary ASD polygon, allocate SV area status and from there a raster of it
  asd.temp = asd[which(asd$dd == i),]
  asd.temp = terra::vect(merge(svareas, asd.temp, by.x = "statevetco", by.y = "svcode"))
  asd.temp <- rasterize(asd.temp, r, field="asdclass")
  
  # Create temporary routes vector
  routestest.temp = terra::vect(routes[which(routes$movementdate == i),])
  
  #Extract the raster value (ASD status) along each point of the route and convert to data.table
  rast.temp = data.table(terra::extract(asd.temp, routestest.temp))
  
  #Loop over the id values for these data and tabulate after allocation of unknown risk to null data - i.e. one route raster per route
  for ( j in unique(sort(rast.temp$ID))){
    subrast.temp = rast.temp[ which(ID == j),]
    # Fill the outputs table with "unknown risk" for empty values
    subrast.temp$asdclass <- ifelse(is.na(subrast.temp$asdclass), "unknown", as.character(subrast.temp$asdclass))
    freq_table = table(subrast.temp$asdclass)
    
    #Calculate the total number of values
    total_values <- sum(freq_table)
    
    #Calculate the proportion of times each risk type appears and append to the prop vectors. If the proportion type doesn't appear in the frequency table, assign it the value 0
    low_prop <- freq_table["low risk"] / total_values
    low_prop = ifelse(is.na(low_prop), 0, low_prop)
    low_propVect = c(low_propVect, low_prop)
    
    high_prop <- freq_table["high risk"] / total_values
    high_prop = ifelse(is.na(high_prop), 0, high_prop)
    high_propVect = c(high_propVect, high_prop)
    
    partial_prop <- freq_table["partial risk"] / total_values
    partial_prop = ifelse(is.na(partial_prop), 0, partial_prop)
    partial_propVect = c(partial_propVect, partial_prop)
    
    unknown_prop <- freq_table["unknown"] / total_values
    unknown_prop = ifelse(is.na(unknown_prop), 0, unknown_prop)
    unknown_propVect = c(unknown_propVect, unknown_prop)
  }
}

#concatenate the proportion vectors together for the violin plot

proportion <- c(low_propVect, high_propVect, partial_propVect, unknown_propVect)
risktype <- rep(c("low risk", "high risk", "partial risk", "unknown"), each=length(low_propVect))

#make a data frame for ggplot
df <- data.frame(risktype, proportion)

ggplot(df, aes(x=risktype, y=proportion)) +
    geom_violin(fill="grey", color="black", draw_quantiles = c(0.5)) +
    theme_classic()+
    xlab("Risk type") +
    ylab("Proportion of route")

# Spread of proportions
dft = data.table(df)

# group the data by "risktype" and get spread group
dft[, .(percentile_95 = quantile(proportion, probs = c(0.025, 0.5, 0.95))), 
                 by = risktype]
```

### Risk proportions for each month of the year

```{r, eval=FALSE}
#Language: R
#Note libraries and base data as section above for annual aggregation

#Empty list to put plots for each month into
plots <- list()

#loop over the months
for (k in 1:12){
  #create sub data frame for routes that took place in month k
  monthroutes <- routes[which(routes$month_number==k),]
  
  #create empty vectors to insert proportion values into
  low_propVect <- c()
  high_propVect <- c()
  partial_propVect <- c()
  unknown_propVect <- c()
  
  #loop over unique movement dates in the month
  for(i in as.character(unique(monthroutes$movementdate))){ # 1. for every movement date 
    #Create temporary ASD polygon, allocate SV area status and from there a raster of it
    asd.temp = asd[which(asd$dd == i),]
    asd.temp = terra::vect(merge(svareas, asd.temp,
               by.x = "statevetco",
               by.y = "svcode"))
    asd.temp <- rasterize(asd.temp, r, field="asdclass")
    
    # Create temporary routes vector
    routestest.temp = terra::vect(
                      monthroutes[which(monthroutes$movementdate == i),])
    
    #Extract the raster value (ASD status) along each point of the route and convert to data.table
    rast.temp = data.table(terra::extract(asd.temp, routestest.temp))
    
    #Loop over the id values for these data and tabulate after allocation of unknown risk to null data - i.e. one route raster per route
    for ( j in unique(sort(rast.temp$ID))){
      subrast.temp = rast.temp[ which(ID == j),]
      # Fill the outputs table with "unknown risk" for empty values
      subrast.temp$asdclass <- ifelse(is.na(subrast.temp$asdclass), "unknown", as.character(subrast.temp$asdclass))
      freq_table = table(subrast.temp$asdclass)
      
      # Calculate the total number of values per rasterised route
      total_values <- sum(freq_table)
      
      # Calculate the proportion of times each risk type appears and append to the prop vectors. If the proportion type doesn't appear in the frequency table, assign it the value 0
      low_prop <- freq_table["low risk"] / total_values
      low_prop = ifelse(is.na(low_prop), 0, low_prop)
      low_propVect = c(low_propVect, low_prop)
      
      high_prop <- freq_table["high risk"] / total_values
      high_prop = ifelse(is.na(high_prop), 0, high_prop)
      high_propVect = c(high_propVect, high_prop)
      
      partial_prop <- freq_table["partial risk"] / total_values
      partial_prop = ifelse(is.na(partial_prop), 0, partial_prop)
      partial_propVect = c(partial_propVect, partial_prop)
      
      unknown_prop <- freq_table["unknown"] / total_values
      unknown_prop = ifelse(is.na(unknown_prop), 0, unknown_prop)
      unknown_propVect = c(unknown_propVect, unknown_prop)
    }
  }
  
  #concatenate the proportion vectors together for the violin plot
  proportion <- c(low_propVect, 
                  high_propVect, 
                  partial_propVect, 
                  unknown_propVect)
  
  risktype <- rep(c("Low", "High", "Part", "Unk"), each=length(low_propVect))
  
  
  #make a data frame for ggplot
  df <- data.frame(risktype, proportion)
  
  #make the violin plot
  p <- ggplot(df, aes(x=risktype, y=proportion)) +
    geom_violin(fill="grey", color="black", draw_quantiles = c(0.5)) +
    theme_classic()+
    ggtitle(paste(k)) +
    xlab("Risk type") +
    ylab("Proportion of route")
  
  #Add the plot to the list of plots
  plots[[k]] <- p
}

#Use grid.arrange to arrange the plots in a grid layout
grid_arranged <- grid.arrange(grobs = plots, nrow = 3 )

grid_arranged
```

## Section 4 - Aggregating the routing data

The aggregation of routing data for the number of movement, average number of equines per movement, average duration and average distance traveled was performed directly in SQL in the database. Below the data is aggregated at month level and unioned to a similar query that is aggregated across the dataset. The final section just creates more readable outcomes for full months rather than month number.

```{sql, eval=FALSE}
--Language: SQL

WITH monthlydata as (
WITH CTE AS (
SELECT EXTRACT(MONTH FROM movementdate::date) as month, totalequids, distance, duration FROM routing2022 
)
SELECT month,
count(*) AS totalmoves,
avg(totalequids) as avgequids,
avg(distance) as avgdist,
avg(duration) as avgdur
FROM CTE
GROUP BY month
ORDER by month),
annualdata as (
WITH CTE AS (
SELECT EXTRACT(MONTH FROM movementdate::date) as month, totalequids, distance, duration FROM routing2022 
)
SELECT 2022 as month,
count(*) AS totalmoves,
avg(totalequids) as avgequids,
avg(distance) as avgdist,
avg(duration) as avgdur
FROM CTE
)
SELECT 
month,
CASE 
WHEN month = 1 THEN 'January'
WHEN month = 2 THEN 'February' 
WHEN month = 3 THEN 'March' 
WHEN month = 4 THEN 'April' 
WHEN month = 5 THEN 'May' 
WHEN month = 6 THEN 'June' 
WHEN month = 7 THEN 'July' 
WHEN month = 8 THEN 'August' 
WHEN month = 9 THEN 'September' 
WHEN month = 10 THEN 'October' 
WHEN month = 11 THEN 'November' 
WHEN month = 12 THEN 'December' 
ELSE 'Other' 
END as periodlabel, 
totalmoves,
round(avgequids, 2) as avgequids, 
round(avgdist::numeric, 2) as avgdist, 
round(avgdur::numeric, 2) as avgdur
FROM monthlydata
UNION ALL
SELECT month, month::text as periodlabel, 
totalmoves,
round(avgequids::numeric, 2) as avgequids, 
round(avgdist::numeric, 2) as avgdis,
round(avgdur::numeric, 2) as avgdur
FROM annualdata
```

### Aggregation of movements per province 

```{r, eval=FALSE}
library(sf)
library(data.table)

points <- read.csv("data/2022Horses.csv")
svareas = st_read("./data/svareas.shp") #State vet areas of South Africa
points_sf <- st_as_sf(points, coords = c("src_longitude", "src_latitude"), crs = st_crs(svareas))

# Join points to polygons
joined <- st_join(points_sf, svareas)

# Convert joined sf object to data.table
joined_dt <- as.data.table(joined)

# Count number of points per polygon using data.table syntax
counts_dt <- joined_dt[, .(n = .N), by = .(province)]
counts_dt[, prop := n / sum(n)]

```

## Section 5 - Transport events for each day of the year

The final section is to generate the Supplementary figure 1 evaluating weekday of movement across 2022.

```{r, eval=FALSE}
#Language: R

library(ggplot2)
library(lubridate)
library(scales)
library(data.table)

data <- read.csv("data/2022Horses.csv")
data$movementdate = as.Date(data$movementdate)

# Group by movementdate and count rows in each group
data = data.table(data)
data[, activity := .N, by = movementdate]
data <- unique(data[, .(movementdate, activity)])

# create week and weekday columns
data$week <- week(data$movementdate)
data$weekday <- lubridate::wday(data$movementdate, label = TRUE)

# create a custom color palette
colors <- c("#ebedf0", "#c6e48b", "#7bc96f", "#239a3b", "#196127")
pal <- scales::manual_pal(colors)

# create the plot
ggplot(data, aes(x = week, y = weekday, fill = activity)) +
  geom_tile(color = "white", width = 1, height = 0.4) +
  scale_fill_gradientn(colors = colors, values = rescale(c(0, max(data$activity))), guide = "none") +
  scale_x_continuous(expand = c(0, 0), position = "top", breaks = seq(1, 52, by = 2)) +
  scale_y_discrete(expand = c(0, 0)) +
  theme_minimal() +
  theme(axis.line = element_blank(), axis.text.x = element_text(size = 9),
        axis.text.y = element_text(size = 9), axis.ticks = element_blank(),
        axis.title = element_blank(),
        panel.grid.major = element_line(color = "white"),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        plot.title = element_text(size = 16, hjust = 0.5, vjust = 2),
        aspect.ratio = 0.5) +
  labs(xlab = "Month of Year") +
  scale_x_continuous(breaks = c(1, 6, 10, 14, 18, 23, 27, 31, 36, 40, 45, 49), 
                     labels = month.abb[c(1:12)]) 
```
